Cosmic web alignments with the shape, angular momentum and peculiar velocities of dark matter halos

### 2014-05-02 paper/main.tex
submitted version
updates de sergio

We study the alignment of dark matter halos with the cosmic web
characterized by the tidal and velocity shear fields. We focus on the
alignment of their shape, angular momentum and peculiar velocities. We
use a cosmological N-body simulation that allows to study dark matter
halos spanning almost five orders of magnitude in mass
($10^{9}$-$10^{14}$) \hMsun and use spatial scales of
$(0.5$-$1.0)$\hMpc to define the cosmic web. After varying the
numerical parameters in our experiments we find that the halo shape
presents the strongest alignment along filaments and walls defined by
the tidal field with a signal that gets stronger as the halo mass
increases; in the case of the velocity shear field only massive halos
$>10^{12}$\hMsun tend to have their shapes perpendicular to filaments
and walls. For the angular momentum we find alignment signals only for
halos more massive than $10^{12}$\hMsun both in the tidal and velocity
shear webs where the preferences are for it to be perpendicular to filaments
or parallel to walls, respectively. Finally, the peculiar velocities
show a strong alignment for all halo masses to move along filaments
and walls in the tidal web, while in the velocity shear the same
alignment is present but weaker and only for halos less massive than
$10^{12}$\hMsun. Most of the different signals between the two web
descriptions are due to the fact that the main directions of stronger
tidal tension and velocity shear coincide for halos below $10^{12}$\hMsun but are
perpendicular to each other for halos more massive than this
threshold.  Our results clearly show that the two different
algorithms we used to define the cosmic web describe different physical
aspects of non-linear collapse and should be used in a complementary
way to understand the effect of the cosmic web on galaxy evolution.
\end{abstract}
\begin{keywords}
methods: N-body simulations, galaxies: haloes, cosmology: theory, dark
matter, large-scale structure of Universe
\end{keywords}

@@

There is a long observational tradition studying galactic properties
as a function of their large scale environment
\citep[e.g.][]{Oemler1974,Dressler1980,Pimbblet2002,Gomez2003,Kauffmann2004,Abbas2006,Baldry2006,Park2007,OMill2008,Gonzales2009,Padilla2010,Wilman2010}. In
these situations the environment definition is usually based on
quantities easily accessible to observations such as local number
density or nearest neighbour measurements \citep{Muldrew2012}. With
the advent of large surveys the visibility of the cosmic web became
clear. As a consequence, in order to capture its filamentarity, the
environment def nition started to be more complex, involving shear
properties of the galaxy density field or the reconstructed dark
matter density field
\citep[e.g.]{Lee2005,Sousbie2008,MunozCuartas2011,Tempel2014}.

@@

In the last decade new algorithms have implemented cosmic web
classifications that go beyond the local density by defining a
location to be a peak, sheet, filament or void depending on the
symmetry properties of the local dark matter distribution. With the
aid of simulations it has been established that halos of a given mass
form earlier in denser regions; concentration, angular momentum and shape can also
be aligned with these web elements. \citep[e.g][]{Hahn2007,Sousbie2008b,Tweb,Gonzalez2010,Noh2011}.

The observational studies that try to measure angular momentum correlations
use the galaxy shape as a proxy  \citep{Lee2002,Lee2007,Jones2010,Paz2008}. In
this respect it is useful to have a theoretical baseline for the
correlations of angular momentum and shape with the cosmic web.  There
is large tradition of alignment measurements of shape and angular momentum
\citep[e.g.][]{Hahn2007,AragonCalvo2007,Zhang2009,Paz2010,Codis2012,Trowland2013,Libeskind2013}. The
main result of these previous studies is that shape alignment is a
robust measurement regardless of the methods and simulations. On the
other hand, the results for the angular momentum differ in the degree
of the alignment.

@@

We caution the reader that the results reported by
\citep{Vweb, Libeskind2013} use a velocity field that is calculated by a
gaussian smoothing of the CIC velocity field without taking into
account any weight by mass. We present a simple example to clarify the difference
between the two ways of finding the smoothed velocity field. Let's consider two
neighboring CIC cells, one with total mass $m_1 = 10^{11}$\hMsun and
velocity along $x$ $v_{1x}=0$ km/s and a second cell with mass $m_2 = 10^{10}$\hMsun and $v_{2x}=-10$ km/s. In the case of defining an average value for the
velocity over these two cells our method finds a result $\bar{v}_x=
(m_1v_{1x} + m_2v_{2x})/(m_1+m_2)= -0.09$km/s while the method used in
\cite{Vweb,Libeskind2013} would find
$\bar{v}_x=(v_{1x}+v_{2x})/2=-5$km/s.

@@

If the halo population does not show any preferential alignment, then
all the halos must be evenly distributed along these three
populations. On the contrary, if there is more than one third of the
halo population in one of these sets, then we can talk about a
preferential alignment along one of the axes. However, this statistics
does not give a precise answer on the degree of the alignment

@@

We now take a look at the angular momentum alignment. Figure
\ref{fig:spin_alignment} shows the results as a function of halo mass
following the same panel layout as in Figure
\ref{fig:shape_alignment}. In all cases we see that these alignment
trends are weaker than the shape alignments. For the Vweb low mass halos
$<10^{12}$\hMsun do not show any preferential alignment with the
cosmic web. Halos more massive than this threshold have their angular
momentum slightly perpendicular to the direction defined by the first
eigenvector and are uncorrelated with the third eigenvector. This translates
into a weak tendency for the angular momentum to lie parallel to walls.

@@

Perhaps the most striking result so far is that the two webs give
different results for the alignment of massive halos. This is not
completely unexpected given that the two algorithms are based on
different physical premises to obtain the directions defining the
eigenvectors. However, we investigate the origin of the different
alignment statistics by studying the inter-web alignment.

@@

A much weaker alignment signal is present for the angular momentum. In
the Tweb only the most massive halos $>10^{12}$\hMsun are anti-aligned
with respect to the filaments, while for the Vweb the massive halos
are aligned with sheets. These results broadly agree with the
published literature. Nevertheless, in some publications
\citep{AragonCalvo2007,Hahn2007} there is an alignment signal
reported at lower halo masses $<10^{12}$\hMsun that we do not detect
in our measurements.

@@

The used the Multi-Scale Morphology Filter to describe th filamentary
structure. The method is based on the Hessian matrix of the density
field, which is computed from the particle distribution using a
Delaunay tesselation field estimatior (DTFE). This allows them to
identify clusters, filaments and walls.

@@

The used the Tweb method applied on three simulations each of $512^3$
particles, with sizes $L_1=45$\hMpc, $L_2=90$\hMpc and
$L_3=180$\hMpc, this corresponds to particle masses of $4.7, 38.0,
300\times 10^7$\hMsun. Halo identification was done with a FOF
algorithm with 0.2 times the interparticle distance. They considered
halos of at least 300  particles.

>>>>

We study the alignment of dark matter halos with the cosmic web
characterized by the tidal and velocity shear fields. We focus on the
alignment of their shape, angular momentum and peculiar velocities. We
use a cosmological N-body simulation that allows to study dark matter
halos spanning almost five orders of magnitude in mass
($10^{9}$-$10^{14}$) \hMsun and use spatial scales of
$(0.5$-$1.0)$\hMpc to define the cosmic web. After varying the
numerical parameters in our experiments we find that the halo shape
presents the strongest alignment along filaments and walls defined by
the tidal field with a signal that gets stronger as the halo mass
increases; in the case of the velocity shear field only massive halos
$>10^{12}$\hMsun tend to have their shapes perpendicular to filaments
and walls. For the angular momentum we find alignment signals only for
halos more massive than $10^{12}$\hMsun both in the tidal and velocity
shear webs where the preferences are for it to be perpendicular to filaments
or parallel to walls, respectively. Finally, the peculiar velocities
show a strong alignment for all halo masses to move along filaments
and walls in the tidal web, while in the velocity shear the same
alignment is present but weaker and only for halos less massive than
$10^{12}$\hMsun. Most of the different signals between the two web
descriptions are due to the fact that the main directions of stronger
tidal tension and velocity shear coincide for halos below $10^{12}$\hMsun but are
perpendicular to each other for halos more massive than this
threshold.  Our results clearly show that the two different
algorithms we used to define the cosmic web describe different physical
aspects of non-linear collapse and should be used in a complementary
way to understand the effect of the cosmic web on galaxy evolution.
\end{abstract}
\begin{keywords}
methods: numerical, galaxies: haloes, cosmology: theory, dark
matter, large-scale structure of Universe
\end{keywords}

@@

There is a long observational tradition studying galactic properties
as a function of their large scale environment
\citep[e.g.][]{Oemler1974,Dressler1980,Pimbblet2002,Gomez2003,Kauffmann2004,Abbas2006,Baldry2006,Park2007,OMill2008,Gonzalez2009,Padilla2010,Wilman2010}. In
these situations the environment definition is usually based on
quantities easily accessible to observations such as local number
density or nearest neighbour measurements \citep{Muldrew2012}. With
the advent of large surveys the visibility of the cosmic web became
clear. As a consequence, in order to capture its filamentarity, the
environment definition started to be more complex, involving shear
properties of the galaxy density field or the reconstructed dark
matter density field
\citep[e.g.]{Lee2005,Sousbie2008,MunozCuartas2011,Tempel2014}.

@@

In the last decade new algorithms have implemented cosmic web
classifications that go beyond the local density by defining a
location to be a peak, sheet, filament or void depending on the
symmetry properties of the local dark matter distribution. With the
aid of simulations it has been established that halos of a given mass
form earlier in denser regions; concentration, angular momentum and shape can also
be aligned with these web elements. \citep[e.g][]{AragonCalvo2007,Hahn2007,Tweb,Zhang2009,Gonzalez2010,Noh2011,Codis2012,Libeskind2013,Trowland2013}.

The observational studies that try to measure angular momentum correlations
use the galaxy shape as a proxy  \citep{Lee2002,Lee2007,Jones2010,Paz2008}. In
this respect it is useful to have a theoretical baseline for the
correlations of angular momentum and shape with the cosmic web.  There
is large tradition of alignment measurements of shape and angular momentum
\citep[e.g.][]{Hahn2007,AragonCalvo2007,Zhang2009,Paz2010,Codis2012,Trowland2013,Libeskind2013,AragonCalvo2014}. The
main result of these previous studies is that shape alignment is a
robust measurement regardless of the methods and simulations. On the
other hand, the results for the angular momentum differ in the degree
of the alignment.

@@

We caution the reader that the results reported by
\citep{Vweb, Libeskind2013,Wang2013} use a velocity field that is calculated by a
gaussian smoothing of the CIC velocity field without taking into
account any weight by mass. 
% We present a simple example to clarify the difference
% between the two ways of finding the smoothed velocity field. Let's consider two
% neighboring CIC cells, one with total mass $m_1 = 10^{11}$\hMsun and
% velocity along $x$ $v_{1x}=0$ km/s and a second cell with mass $m_2 = 10^{10}$\hMsun and $v_{2x}=-10$ km/s. In the case of defining an average value for the
% velocity over these two cells our method finds a result $\bar{v}_x=
% (m_1v_{1x} + m_2v_{2x})/(m_1+m_2)= -0.09$km/s while the method used in
% \cite{Vweb,Libeskind2013} would find
% $\bar{v}_x=(v_{1x}+v_{2x})/2=-5$km/s.

@@

If the halo population does not show any preferential alignment, then
all the halos must be evenly distributed along these three
populations. On the contrary, if there is more than one third of the
halo population in one of these sets, then this will indicate a
preferential alignment along one of the axes. However, this statistics
does not give a precise answer on the degree of the alignment

@@

We now focus our attention on the angular momentum alignment. Figure
\ref{fig:spin_alignment} shows the results as a function of halo mass
following the same panel layout as in Figure
\ref{fig:shape_alignment}. In all cases we see that these alignment
trends are weaker than the shape alignments. For the Vweb low mass halos
$<10^{12}$\hMsun do not show any preferential alignment with the
cosmic web. Halos more massive than this threshold have their angular
momentum slightly perpendicular to the direction defined by the first
eigenvector and are uncorrelated with the third eigenvector. This translates
into a weak tendency for the angular momentum to lie parallel to walls.

@@

Perhaps the most striking result so far is that the two webs algorithms gives
different results for the alignment of massive halos. This is not
completely unexpected given that the two algorithms are based on
different physical premises to obtain the directions defining the
eigenvectors. However, we investigate the origin of the different
alignment statistics by studying the inter-web alignment.

@@

A much weaker alignment signal is present for the angular momentum. In
the Tweb only the most massive halos $>10^{12}$\hMsun are anti-aligned
with respect to the filaments, while for the Vweb the massive halos
are aligned with sheets. These results broadly agree with the
published literature. Nevertheless, in some publications
\citep{AragonCalvo2007,Hahn2007,AragonCalvo2014} there is an alignment signal
reported at lower halo masses $<10^{12}$\hMsun that we do not detect
in our measurements.

@@

They used the Multi-Scale Morphology Filter to describe th filamentary
structure. The method is based on the Hessian matrix of the density
field, which is computed from the particle distribution using a
Delaunay tesselation field estimatior (DTFE). This allows them to
identify clusters, filaments and walls.

@@

They used the Tweb method\footnote{In \citep{Hahn2007} the authors do not use the term "Tweb", but their definition of the
tidal field tensor they use is equivalent with the T-Web} applied on three simulations each of $512^3$
particles, with sizes $L_1=45$\hMpc, $L_2=90$\hMpc and
$L_3=180$\hMpc, this corresponds to particle masses of $4.7, 38.0,
300\times 10^7$\hMsun. Halo identification was done with a FOF
algorithm with 0.2 times the interparticle distance. They considered
halos of at least 300  particles.
